---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r function section, echo=FALSE,include=FALSE}

library(Seurat)
library(tidyverse)
library(ggrepel)

# Well known marker genes of airway epithelial, mesenchymal and immune cells
markers <- c("MKI67", "TOP2A", "CDC20",     # Cycling cells
             "KRT5", "KRT15", "TP63",       # Basal cells
             "SPRR3", "SERPINB2", "KRT13",  # Suprabasal cells
             "SCGB1A1", "VMO1", "BPIFA1",   # Secretory cells
             "MUC5AC", "MUC5B", "CXCL3",    # Goblet cells,
             "CDC20Bshort", "CCNO", "HES6", # Deuterosomal cells
             "FOXJ1", "TPPP3", "PIFO",      # Multiciliated cells
             "CFTR", "ASCL3", "ATP6V1B1",   # Ionocytes
             "DARC", "VIM", "SPARCL1",      # Endothelial cells
             "CD68", "TYROBP", "LST1",      # Macrophages
             "TPSAB1","SLC18A2","CPA3",     # Mast cells
             "NKG7","GZMA","CD2",           # T cells
             "CD79A","MS4A1","CD79B",       # Plasma cells
             "LTF","TCN1","LYZ",            # Serous cells
             "TAGLN", "ACTA2", "MYL9",      # Smooth muscle
             "FBLN1", "DPT", "PTGDS",       # Fibroblast
             "GIMAP7", "CD8A", "PTPN7"      # B cells,
             ) 

markers_cluster <- c(rep("Cycling cells", 3), rep("Basal", 3),
                     rep("Suprabasal", 3), rep("Secretory", 3),
                     rep("Goblet", 3), rep("Deuterosomal", 3),
                     rep("Multiciliated", 3), rep("Ionocyte", 3),
                     rep("Endothelial", 3), rep("Macrophages", 3),
                     rep("Mast cell", 3), rep("T cells", 3),
                     rep("Plasma cells", 3), rep("Serous cells", 3), rep("Smooth Muscle", 3),
                     rep("Fibroblast", 3), rep("B cells", 3))

cluster_color <- read.csv(file = "/data/deprez_data/HCA/Analysis/FullDataset_v6/HCA_airway_epithelium_v6/clusterColor.tsv",
                          sep = "\t", stringsAsFactors = F, header = T)
cell_type_color <- cluster_color$color
names(cell_type_color) <- cluster_color$Cluster
```

---
title: "Analyse 1 - D344_Biop_Int1"
author: "Marie Deprez"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
---

# Primary exploratory analysis

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=6, fig.width=11}
setwd("~/HCA/")

# Load dataset directly from 10X output repository
data <- Read10X("DataReal/D344_Biop_Int1/outs/filtered_gene_bc_matrices/ucagenomix-cellranger-hg19-1.3.0/")
D344YI1 <- CreateSeuratObject(raw.data = data, min.cells = 3)
dim(D344YI1@data)


# Identify data quality metrics
mito.genes <- grep("^MT-", rownames(as.matrix(D344YI1@raw.data)), value = T)
rpsrpl.genes <- grep("^RP[SL]", rownames(as.matrix(D344YI1@raw.data)), value = T)

dropouts <- Matrix::colSums(D344YI1@data == 0)/nrow(D344YI1@raw.data)
percent.mito <- Matrix::colSums(D344YI1@raw.data[mito.genes, ])/Matrix::colSums(D344YI1@raw.data)
percent.rpsrpl <- Matrix::colSums(D344YI1@raw.data[rpsrpl.genes, ])/Matrix::colSums(D344YI1@raw.data)

D344YI1 <- AddMetaData(object = D344YI1, metadata = dropouts, col.name = "dropouts")
D344YI1 <- AddMetaData(object = D344YI1, metadata = percent.mito, col.name = "percent.mito")
D344YI1 <- AddMetaData(object = D344YI1, metadata = percent.rpsrpl, col.name = "percent.rps.rpl")

VlnPlot(object = D344YI1, features.plot = c("nGene", "nUMI", "percent.mito", 
                                            "percent.rps.rpl", "dropouts"), 
        nCol = 5, cols.use = "lightsteelblue3", point.size.use = 0.5) 
```

### Filter cells according to their quality metrics

Parameter        | Min threshold    | Max threshold
---------------- | ---------------- | ---------------
nGene            | 400              | +Inf
nUMI             | -Inf             | 20000
Percent.mito     | -Inf             | 0.1
Dropouts         | -Inf             | 0.97 


```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=6, fig.width=11}
D344YI1 <- FilterCells(object = D344YI1, subset.names = c("nGene", "nUMI", 
                                                          "percent.mito", "dropouts"), 
                              low.thresholds = c(400,-Inf, -Inf, -Inf), 
                              high.thresholds = c(+Inf, 20000, 0.1, 0.97))

# Check quality metrics distribution following cell filtering
VlnPlot(object = D344YI1, features.plot = c("nGene", "nUMI", "percent.mito", 
                                            "percent.rps.rpl", "dropouts"), 
        nCol = 5, cols.use = "lightsteelblue3", point.size.use = 0.5) 
```

Before  **`r dim(D344YI1@raw.data)[2]`** cells   -->   After  **`r dim(D344YI1@data)[2]`** cells

***

### Identification of highly variable genes (HVGs) and dimension reduction (PCA)

Median UMI per cell : **`r median(D344YI1@meta.data$nUMI)`**

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=5, fig.width=5}
D344YI1 <- NormalizeData(object = D344YI1, normalization.method = "LogNormalize", 
                         scale.factor = median(D344YI1@meta.data$nUMI))
D344YI1 <- FindVariableGenes(object = D344YI1, mean.function = ExpMean, 
                             dispersion.function = LogVMR, 
                             x.high.cutoff = 3, x.low.cutoff = 0.03, y.cutoff = 0.2)
```

Nb highly variable genes : **`r length(D344YI1@var.genes)`**

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=10, fig.width=10}
D344YI1 <- ScaleData(D344YI1, genes.use=D344YI1@var.genes, 
                     model.use = "negbinom", display.progress = FALSE)

D344YI1 <- RunPCA(D344YI1, pcs.compute = 30, weight.by.var = FALSE, display.progress = FALSE, do.print=FALSE)

# Compute correlation score between each gene and the components to first identify markers of heterogeneity.
D344YI1 <- ProjectPCA(object = D344YI1, do.print = FALSE)

PCHeatmap(object = D344YI1, pc.use = 1:9, cells.use = 500, do.balanced = TRUE, label.columns = FALSE)
PCHeatmap(object = D344YI1, pc.use = 10:18, cells.use = 500, do.balanced = TRUE, 
          label.columns = FALSE, use.full = FALSE)
```

Nb PCs to use : **PCs : 1:10**

#### Dataset embbedings (t-SNE and UMAP) and clustering (Louvain algorithm)

```{r message=FALSE, warning=FALSE, message = F, echo=FALSE, eval=T, fig.height=4, fig.width=9}
D344YI1 <- RunTSNE(D344YI1, dims.use = c(1:10), do.fast = T)
D344YI1<- FindClusters(D344YI1, resolution=0.4, reduction.type = "pca", k.param = 10, 
                       dims.use = c(1:10), save.SNN = T, force.recalc = T)

p1 <- TSNEPlot(D344YI1, do.return = TRUE, do.label = TRUE, pt.size=2 )

D344YI1 <- RunUMAP(D344YI1, dims.use = c(1:10))
p2 <- DimPlot(object = D344YI1, reduction.use = 'umap', do.label = T, 
              pt.size=2, do.return = T, no.legend = T)

plot_grid(p1,p2)
```

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=6, fig.width=6}
FeaturePlot(D344YI1, c("nUMI", "nGene", "percent.mito", "percent.rps.rpl"), nCol = 2,
            cols.use = c("grey70", "red4"))
```

Distribution of the quality metrics per cluster

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=4, fig.width=8}
VlnPlot(object = D344YI1, features.plot = c("nUMI", "nGene"), nCol = 2,
        x.lab.rot = 0.5, point.size.use = 0.5)
VlnPlot(object = D344YI1, features.plot = c("percent.mito", "percent.rps.rpl"), nCol = 2,
        x.lab.rot = 0.5, point.size.use = 0.5)
```

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=4, fig.width=4}
GenePlot(object = D344YI1, gene1 = "nUMI", gene2 = "nGene", pch.use=20)
```

#### Identification of cluster marker genes (Wilcoxon's rank test)

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=12, fig.width=11}
D344YI1.markers <- FindAllMarkers(object = D344YI1, only.pos = TRUE, min.pct = 0.5, thresh.use = 0.5)
top5 <- D344YI1.markers %>% group_by(cluster) %>% top_n(5, avg_logFC)
DoHeatmap(object = D344YI1, genes.use=top5$gene, slim.col.label = TRUE, 
          remove.key = TRUE, cex.row = 8, cex.col = 8,
          col.low = "#0033cc", col.mid = "#e6e6e6", col.high = "#990000")
```


***
***

###### Cell type labelling

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=4, fig.width=9}
current.cluster.ids <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9 , 10, 11)
new.cluster.ids<- c("Basal", "Endothelial", "Suprabasal", "Endothelial",
                    "Smooth muscle", "Secretory",  "Smooth muscle", "Fibroblast",
                    "Macrophage", "Glandulaire", "Multiciliated", "Immune")

D344YI1@ident <- plyr::mapvalues(x = D344YI1@ident, from = current.cluster.ids, to = new.cluster.ids)

table(D344YI1@ident)


p1 <- TSNEPlot(object = D344YI1, do.label = F, no.legend = F, pt.size = 2, 
         do.return = T, colors.use = colors.BBInt) 
p2 <- DimPlot(object = D344YI1, reduction.use = 'umap', do.label = F, pt.size=2, do.return = T,
          cols.use = colors.BBInt, no.legend = T)
plot_grid(p1,p2)
```

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=6, fig.width=9}
markers_flt <- markers[markers %in% rownames(D344YI1@data)]
DotPlot(D344YI1, genes.plot = markers_flt, x.lab.rot = 45)
```


```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=5, fig.width=9}
# Focus on some genes of interest
FeaturePlot(D344YI1, features.plot = c("GRP", "SCG5", "PCSK1N",
                                       "CDC20Bshort", "CCNO", "HES6"),nCol = 3,
            cols.use = c("lightgrey","red"))
```

Recompute top marker genes for each cell population
```{r message=FALSE, warning=FALSE, echo=FALSE, eval=T, fig.height=16, fig.width=14}
D344YI1.markers <- FindAllMarkers(object = D344YI1, only.pos = TRUE, min.pct = 0.25, thresh.use = 0.25)
top5 <- D344YI1.markers %>% group_by(cluster) %>% top_n(10, avg_logFC)
DoHeatmap(object = D344YI1, genes.use=top5$gene, slim.col.label = TRUE, 
          remove.key = TRUE, cex.row = 8, cex.col = 8, group.label.rot = T,
          col.low = "#0033cc", col.mid = "#e6e6e6", col.high = "#990000")

# Format the top table accordingly to each cell cluster
marker_k0 <- D344YI1.markers %>% filter(cluster=="Basal") %>% arrange(desc(avg_logFC)) %>% mutate(ratio = pct.1 / pct.2) %>% select(cluster,gene,avg_logFC,p_val_adj,pct.1,pct.2,ratio)
marker_k1 <- D344YI1.markers %>% filter(cluster=="Suprabasal") %>% arrange(desc(avg_logFC)) %>% mutate(ratio = pct.1 / pct.2) %>% select(cluster,gene,avg_logFC,p_val_adj,pct.1,pct.2,ratio)
marker_k2 <- D344YI1.markers %>% filter(cluster=="Secretory") %>% arrange(desc(avg_logFC)) %>% mutate(ratio = pct.1 / pct.2) %>% select(cluster,gene,avg_logFC,p_val_adj,pct.1,pct.2,ratio)
marker_k3 <- D344YI1.markers %>% filter(cluster=="Endothelial") %>% arrange(desc(avg_logFC)) %>% mutate(ratio = pct.1 / pct.2) %>% select(cluster,gene,avg_logFC,p_val_adj,pct.1,pct.2,ratio)
marker_k4 <- D344YI1.markers %>% filter(cluster=="Fibroblast") %>% arrange(desc(avg_logFC)) %>% mutate(ratio = pct.1 / pct.2) %>% select(cluster,gene,avg_logFC,p_val_adj,pct.1,pct.2,ratio)
marker_k5 <- D344YI1.markers %>% filter(cluster=="Smooth Muscle") %>% arrange(desc(avg_logFC)) %>% mutate(ratio = pct.1 / pct.2) %>% select(cluster,gene,avg_logFC,p_val_adj,pct.1,pct.2,ratio)
marker_k6 <- D344YI1.markers %>% filter(cluster=="Immune") %>% arrange(desc(avg_logFC)) %>% mutate(ratio = pct.1 / pct.2) %>% select(cluster,gene,avg_logFC,p_val_adj,pct.1,pct.2,ratio)
marker_k7 <- D344YI1.markers %>% filter(cluster=="Serous") %>% arrange(desc(avg_logFC)) %>% mutate(ratio = pct.1 / pct.2) %>% select(cluster,gene,avg_logFC,p_val_adj,pct.1,pct.2,ratio)
marker_k8 <- D344YI1.markers %>% filter(cluster=="Multiciliated") %>% arrange(desc(avg_logFC)) %>% mutate(ratio = pct.1 / pct.2) %>% select(cluster,gene,avg_logFC,p_val_adj,pct.1,pct.2,ratio)
marker_k9 <- D344YI1.markers %>% filter(cluster=="Macrophage") %>% arrange(desc(avg_logFC)) %>% mutate(ratio = pct.1 / pct.2) %>% select(cluster,gene,avg_logFC,p_val_adj,pct.1,pct.2,ratio)

all_marker_list <- bind_rows(marker_k0,marker_k1,marker_k2,marker_k3,
                             marker_k4,marker_k5,marker_k6,
                             marker_k7,marker_k8,marker_k9,.id = NULL)

# Save the marker table and annotated Seurat object.
#write.table(all_marker_list,
            #file = "/home/deprez/HCA/DataReal/D344_Biop_Int1/all_markers_D344YI1.xls",
            #sep = "\t",row.names = F)
#save(D344YI1, file = "/home/deprez/HCA/DataReal/D344_Biop_Int1/D344YI1.Rda")
```




